Two applications of logic programming to Coq

### 2021-03-05 papers/types2020/revisions/main.tex
cleanup of addition of conversion

\begin{abstract}
The logic programming paradigm provides a flexible setting for
representing, manipulating, checking, and elaborating proof
structures.  This is particularly true when the logic programming
language allows for bindings in terms and proofs.  In this paper,
we make use of two recent innovations at the intersection of logic
programming and proof checking.  One of these is the
\emph{foundational proof certificate} (FPC) framework which provides a
flexible means of defining the semantics of a range of proof
structures for classical and intuitionistic logic.  A second
innovation is the recently released Coq-Elpi plugin for Coq in which
the Elpi implementation of \lP can send and retrieve information to
and from the Coq kernel.  We illustrate the use of both this Coq
plugin and FPCs with two example applications.  First, we implement an
FPC-driven sequent calculus for a fragment of the Calculus of
Inductive Constructions and we package it into a tactic to perform
property-based testing of inductive types corresponding to pure
(higher-order) Horn programs.  Second, we implement in Elpi a proof
checker for first-order intuitionistic logic and demonstrate how proof
certificates can be supplied by external (to Coq) provers and then
elaborated into the fully detailed proof terms that can be checked by
the Coq kernel.
\end{abstract}

>>>>

\begin{abstract}
  The logic programming paradigm provides a flexible setting for
  representing, manipulating, checking, and elaborating proof
  structures.  This is particularly true when the logic programming
  language allows for bindings in terms and proofs.  In this paper, we
  make use of two recent innovations at the intersection of logic
  programming and proof checking.  One of these is the
  \emph{foundational proof certificate} (FPC) framework which provides
  a flexible means of defining the semantics of a range of proof
  structures for classical and intuitionistic logic.  A second
  innovation is the recently released Coq-Elpi plugin for Coq in which
  the Elpi implementation of \lP can send and retrieve information to
  and from the Coq kernel.  We illustrate the use of both this Coq
  plugin and FPCs with two example applications.  First, we implement
  an FPC-driven sequent calculus for a fragment of the Calculus of
  Inductive Constructions and we package it into a tactic to perform
  property-based testing of inductive types corresponding to Horn
  clauses.  Second, we implement in Elpi a proof checker for
  first-order intuitionistic logic and demonstrate how proof
  certificates can be supplied by external (to Coq) provers and then
  elaborated into the fully detailed proof terms that can be checked
  by the Coq kernel.
\end{abstract}

### 2021-03-05 papers/types2020/revisions/sec3.tex
cleanup of addition of conversion

\begin{figure}[t]
\centering
  \[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A) : T \in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  A : s}{E[] \vdash_{CIC} A : s\quad \sortExpert{\XXi}}% The order should be s : A, not A : s ?
  %% we are checking nat : Set, for example, so A : s --AM
  \qquad
  \infer{\bc{\XXi}{}{[]} A A'}{E[]\vdash_{CIC} A =_{\beta\delta\iota\xi}A'\quad\initLExpert{\XXi}}
\]
  \vskip -18pt
\[
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\quad
         \bc{\XXi_2}{}l {D[t/x]} A\quad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of the checking of a proof certificate and the
  synthesis of a dependently typed $\lambda$-term.}
\label{fig:augmented}
\end{figure}

\begin{figure}
\lstinputlisting[language=lprolog,linerange={sigs-end}]{../code/dep-kernel.sig}
\lstinputlisting[language=lprolog,linerange={check-end}]{../code/dep-kernel-v2.mod}
\caption{Implementation of a  dependent kernel for Coq terms.}
\label{fig:kernel}
\end{figure}

Figure~\ref{fig:augmented} contains these proof rules and extends them
in the following ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We account for \emph{computation} inside atoms by generalizing
    to \emph{init} rule to work modulo conversion.
  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
  \item The inference rules are  annotated by terms structures that
    can be given directly to the Coq kernel for checking.
  \item We have added various premises which are responsible for
    pulling information across  Coq's global environment.
\end{enumerate}
The proof system described in Figure~\ref{fig:augmented} (and
implemented in Figure~\ref{fig:kernel}) corresponds to a subset of the Calculus
of Inductive Constructions in which inductive definitions are
limited to first-order Horn clauses.
%
This  system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories.
% %
% In our case, the \lP goal formula \lsti{(check Xi (go B T))}
% corresponds to $\XXi\vdash B \colon T$ and the formula
% %
% \lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
% D\vdash B$.
%
This style of  proof terms coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

The proof system is parameterized by Coq's global environment $E$,
here a set of constant and inductive definitions; following Coq's
reference manual, we denote such a definition with
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the (possibly mutually) inductive type and
$\Gamma_C$ the names and types of its constructors; finally $p$
denotes the number of parameters and plays here no role. We delegate to
Coq's type checking the enforcement of the well-sortedness of
inductive types. The \emph{local context} is empty, since we are only
dealing with types that correspond to Horn clauses, and atomic types
are inductively defined. In fact, we do not have a $\forall$ rule on
the right, although the proof theory would gladly allow it. This means
at the term level that there are no variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors of an \lsti{Inductive} definition from the global
environment, one that matches the head symbol of the atom we aim to backchain on, and then
call the latter judgment that will find a correct instantiation, if any. The rules for backchaining include
 the (conflation of the) left introduction rules for
$\forall$ and $\supset$, as well as the \emph{init} rule, which incorporates Coq's conversion.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants that we use:
\begin{lstlisting}[language=lprolog]
kind term   type.                                 % reification of Coq terms
kind gref   type.                                 % reif. of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products by taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
implementation of the product-left rule, we apply \lsti{D} to the variable
\lsti{Tm}, we get a new term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in the  LJT and PTSC's tradition; this is consistent with proof search in
our application being cut-free, whereas explicit substitutions are linked to cuts.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to focus on from the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase.

@@

The following
example shows how we can use the above tactic to do FPC-driven logic
programming in Coq and return a Coq proof-term:
\begin{lstlisting}[]
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 ([0] $++$ [1]) R.
elpi dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 ([0] $++$ [1]) R)   [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
\end{lstlisting}

The \lsti{dprolog} tactic can be seen as a programmable (and possibly
faster) version of Coq's \lsti{eauto}% , although it currently lacks the
% latter's capability of solving goals by
% \lsti{reflexivity}
; % as well as the modularity of using
% \lsti{Hints} as introduction rules
 \lsti{dprolog} is not restricted to depth-first search, since
it follows the dictate of the given FPC; for example we could easily add
iterative-deepening search. Furthermore, the tactic can still use
\lsti{Hints} via Elpi's \lsti{Db} feature, and FPCs can provide a
\emph{trace} that may be more customizable than the one offered by 
\lsti{(e)auto}'s hard-wired \lsti{Debug} facility.

>>>>

\begin{figure}[t]
\centering
  \[
\infer{\XXi_1\vdash k\ l : A}
      { \mathrm{Ind}[p] \ (\Gamma_I := \Gamma_C) \in E\quad 
       (\mathit{head} \ A) : T \in \Gamma_I\quad  
       k : D \in \Gamma_C\quad 
       \bc{\XXi_2}{} l D A \quad\unfoldExpert{\XXi_1}{\XXi_2}} 
\]
\vskip -18pt
\[
  \infer{\XXi\vdash  A : s}{E[] \vdash_{CIC} A : s\quad \sortExpert{\XXi}}% The order should be s : A, not A : s ?
  %% we are checking nat : Set, for example, so A : s --AM
  \qquad
  \infer{\bc{\XXi}{}{[]} A A'}{E[]\vdash_{CIC} A =_{\beta\delta\iota\xi}A'\quad\initLExpert{\XXi}}
\]
  \vskip -18pt
\[
  \infer{\bc{\XXi}{}{(t\ ::\ l)}{\depprod{x}{B}{D}}{A}}
        {\XXi_1\vdash t : B\quad
         \bc{\XXi_2}{}l {D[t/x]} A\quad
         \prodE{\XXi}{\XXi_1}{\XXi_2}{t}}
\]
\caption{Specification of a core calculus% of the checking of a proof certificate and the
  % synthesis of a dependently typed $\lambda$-term
  .}
\label{fig:augmented}
\end{figure}

\begin{figure}
\lstinputlisting[language=lprolog,linerange={sigs-end}]{../code/dep-kernel.sig}
\lstinputlisting[language=lprolog,linerange={check-end}]{../code/dep-kernel.mod}
\caption{Implementation of the core calculus.}
\label{fig:kernel}
\end{figure}

Figure~\ref{fig:augmented} contains these proof rules and extends them
in the following ways so that it more appropriately fits into
the Coq proof system.
\begin{enumerate}
  \item Instead of having separate connectives for $\forall$ and
    $\supset$, we have the dependent product connective $(x:A)D$. 
  \item We account for \emph{computation} inside atoms by generalizing
    the \emph{init} rule to work modulo conversion.
  \item We have incorporated both \emph{proof certificates}~\cite{chihani17jar} (using the
    schematic variable $\XXi$) along with expert predicates
    (predicates with the $e$ subscript). We explain this in Section~\ref{ssec:cert}.
  \item The inference rules are  annotated by terms structures that
    can be given directly to the Coq kernel for checking.
  \item We have added various premises which are responsible for
    pulling information across  Coq's global environment.
\end{enumerate}
The proof system described in Figure~\ref{fig:augmented} (and
implemented in Figure~\ref{fig:kernel}) corresponds to a subset of the Calculus
of Inductive Constructions in which inductive definitions are
limited to % first-order
Horn clauses.
%
This  system is inspired by the calculus for proof search in
Pure Type Systems introduced in \cite{LengrandDM06}, based in turn on
ideas stemming from focusing (in particular, the LJT calculus of
Herbelin \cite{Herbelin94}). Similar to that calculus, we have a
term language that includes terms and lists of terms, and two distinct
typing judgments for the two categories.
% %
% In our case, the \lP goal formula \lsti{(check Xi (go B T))}
% corresponds to $\XXi\vdash B \colon T$ and the formula
% %
% \lsti{(check Xi (bc D B L))} corresponds to $\XXi\Downarrow L\colon
% D\vdash B$.
%
This style of  proof terms coincides with the idea behind the \emph{spine
  calculus}~\cite{Cervesato97tr}.  The main novelty of our proof
system here is that proof terms and proof certificates are used
simultaneously in all inference rules.

The proof system is parameterized by Coq's global environment $E$,
here a set of constant and inductive definitions; following Coq's
reference manual, inductive definition are denoted by
$\mathrm{Ind}[p] (\Gamma_I := \Gamma_C)$, where $\Gamma_I$ determines
the names and types of the (possibly mutually) inductive type and
$\Gamma_C$ the names and types of its constructors; finally $p$
denotes the number of parameters and plays here no role. We delegate to
Coq's type checking the enforcement of the well-sortedness of
inductive types. The \emph{local context} is empty, since we are only
dealing with types that correspond to Horn clauses, and atomic types
are inductively defined. In fact, we do not have a $\forall$ rule on
the right, although the proof theory would gladly allow it. This means
at the term level that there are no variables in our grammar of
terms. Terms are always applied to a (possibly empty) list of
arguments.  The \emph{decide} rule, % is the equivalent in our
% a \emph{decide} rule in an explicitly focused system
% \cite{liang09tcs}.  
as in the previous proof system for Horn logic,  given an atom,
 selects a clause on which to back-chain on: we lookup the
constructors of an inductive definition from the global
environment, one that matches the head symbol of the atom we aim to backchain on, and then
call the latter judgment that will find a correct instantiation, if any. The rules for backchaining include
 the (conflation of the) left introduction rules for
$\forall$ and $\supset$, as well as the \emph{init} rule, which incorporates Coq's conversion.

@@

\subsubsection{Coq-specific code}
Coq terms are accessed through the Coq-Elpi API, and their representation
in \lP takes advantage of native \lP constructs such as lists and binders.
The following is part of the Coq-Elpi API signature of constants that we use:
\begin{lstlisting}[language=lprolog]
kind term   type.                                 % reification of Coq terms
kind gref   type.                                 % reif. of references to global terms
type global gref -> term.                            % coercion to term
type indt   inductive -> gref.                       % reification of inductive types
type indc   constructor -> gref.                     % reification of their constructors 
type app    list term -> term.                       % reification of nary application
type prod   name -> term -> (term -> term) -> term. % reification of dependent product
\end{lstlisting}
%
Note that \lsti{prod} encodes dependent products by taking a name for pretty
printing, a term and a \lP abstraction from terms to terms: i.e., $(x :
B) D$ is encoded by \verb|prod ``x'' B (x\ D x)|; when, in the
implementation of the product-left rule, we apply \lsti{D} to the variable
\lsti{Tm}, we get a new term that can be used to continue backchaining. This
application is obtained via meta-level substitution, in the style of HOAS. In
this sense, our calculus uses \emph{implicit} substitutions, rather than
explicit ones as in the  LJT and PTSC's tradition; this is consistent with proof search in
our application being cut-free, whereas explicit substitutions are linked to cuts.
% Syntax: one can think as if something of type \lsti{term -> term} were a
% \lsti{term} with a hole in it for another \lsti{term}.
The decide rule makes use of the Coq-Elpi primitives
\lsti{coq.safe-dest-app} to obtain the head term of a (possibly
nested) application, and \lsti{coq.env.indt} to access the global
environment of inductive definitions and query for information about them.
The \lsti{decideE} predicate,  has, among others, the role of selecting which
constructor to focus on from the inductive type. The kernel will successively
obtain the type of the selected constructor, and initiate the backchaining
phase. The latter is terminated when the focused atom unifies with the current goal, via
Elpi's primitive \lsti{coq.unify-eq}.

@@

The following
example shows how we can use the above tactic to do FPC-driven logic
programming modulo conversion in Coq and return a Coq proof-term:
\begin{lstlisting}[]
Inductive insert (x:nat) : list nat -> list nat -> Prop :=
| i_n : insert x [] [x]
| i_s : forall y: nat, forall ys, x <= y -> insert x (y :: ys) (x :: y :: ys)
| i_c : forall y: nat, forall ys rs, y <= x -> insert x ys rs -> insert x (y :: ys) (x :: rs).
Lemma i1:  exists R, insert 2 ([0] $++$ [1]) R.
elpi dprolog height 10.
Qed.
Print i1.
ex_intro (fun R : list nat => insert 2 ([0] $++$ [1]) R)   [0; 1; 2]
  (i_c 2 0 [1] [1; 2] (le_S 0 1 (le_S 0 0 (le_n 0)))
	 (i_c 2 1 [] [2] (le_S 1 1 (le_n 1)) (i_n 2)))
\end{lstlisting}

The \lsti{dprolog} tactic can be seen as a programmable (and possibly
faster) version of Coq's \lsti{eauto}% , although it currently lacks the
% latter's capability of solving goals by
% \lsti{reflexivity}
; % as well as the modularity of using
% \lsti{Hints} as introduction rules
 \lsti{dprolog} is not restricted to depth-first search, since
it follows the dictate of the given FPC; for example we could easily add
iterative-deepening search. Furthermore,
%% commented out till we know for sure:
%%the tactic can still use\lsti{Hints} via Elpi's \lsti{Db} feature, and
 FPCs can provide a
\emph{trace} that may be more customizable than the one offered by 
\texttt{(e)auto}'s hard-wired \texttt{Debug} facility.
